Linear analysis of the vertical shear instability in astrophysical disks

### 2015-02-07 appendix.tex
undergoing major revisions

\section{Local model for a global problem}\label{global_corr}
Here, we discuss the relation between the linear problem in the
VGSB framework and the global linear problem. We begin with the
axisymmetric linearized fluid equations in cylindrical co-ordinates, including
thermal relaxation, 
\begin{align}
\ii\sigma\delta P &=\frac{\p P}{\p z}\delta v_z + \gamma P
\left[\frac{1}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p
    \delta v_z}{\p z}\right] + \frac{1}{t_c}\left(\delta P -
  \frac{P}{\rho}\delta\rho\right) + \hat{g}_c \frac{\p P}{\p r}\delta v_r,\\
  \ii\sigma\delta\rho &= \frac{\rho}{r}\frac{\p}{\p r}\left(r\delta v_r\right) + \frac{\p}{\p
  z}\left(\rho\delta v_z\right)  + \hat{g}_c\frac{\p\rho}{\p r}\delta v_r,\\
\ii\sigma\delta v_r &= \frac{1}{\rho}\frac{\p\delta P}{\p r} - 2\Omega
\delta v_\phi - \hat{g}_c\frac{\delta\rho}{\rho^2}\frac{\p P}{\p r},\\ 
\ii\sigma\delta v_\phi  &= \frac{\kappa^2}{2\Omega}\delta v_r +
r\frac{\p\Omega}{\p z}\delta v_z,\\
\ii \sigma\delta v_z &= \frac{1}{\rho}\frac{\p\delta P}{\p z} -
\frac{\delta \rho}{\rho^2}\frac{\p P}{\p z},
\end{align}
where the coefficients $\hat{g}_c=1$ are inserted to label terms
associated with the background radial disk structure.

We assume perturbations have the form $W(z)\exp{\ii k_r r}$ such
that $|k_rr|\gg1$, so that curvature terms can be ignored. For these
small radial lengthscale perturbations we also approximate the
background structure and radial gradients as uniform in $r$, and
evaluate all cofficients in the above equations at the fiducial radius
$r=r_0$. This procedure produces $\phi$ and $z$ momentum equations
that are equilvalent to those for $y$ and $z$ momenta in the VGSB.

@@

>>>>

\section{Limitations of the vertically global shearing box}\label{global_corr}
Here, we discuss the relation between the linear problem in the
VGSB framework and the global linear problem.

 This procedure produces $\phi$ and $z$ momentum equations
that are equilvalent to those for $y$ and $z$ momenta in the VGSB.

@@

\subsection{Spurious growth of adiabatic perturbations in the local
  model with vertical shear}\label{analytic_adia} 
For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
simply use the Soilberg-Hoiland critera to assess stability
(\S\ref{solberg}). Nevertheless, we briefly consider this case in the
local framework to demonstrate its limitations. For $\beta\to\infty$
the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
be written as

\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat \notag\\
  &=  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
  &\phantom{=} +
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat\notag\\
  &\phantom{=} + \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}) and is a result of the
radially-local approximation (Appendix \ref{global_corr}). To 
avoid this, we must restrict our attention to short thermal relaxation
timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
($\gamma=\Gamma$) when $q\neq0$.

### 2015-02-07 linear_adia.tex
undergoing major revisions

Using the expressions for $\delta\chi$ and $\hat{\sigma}$ we can solve
for $\delta\hat{\sigma}$. This gives 
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
      q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by finite cooling is ineffective at very large or very
small scales. We plot Eq. \ref{nearly_iso_dsig} in Fig. \ref{domegadbeta} with
$\epsilon=0.05$, $q=-1.0$ and $\gamma=1.4$. For this disk model the
optimum wavenumber for stabilization is $\hat{k}\sim 5$---$6$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/domegadbeta}
  \caption{Rate of change of the fundamental VSI eigenfrequency
    $\sigma$ with respect to the introduction of a small but finite
    cooling time. The plotted quantity is $\p\hat{\sigma}/\p\beta$ at $\beta=0$, evaluated using
    Eq. \ref{nearly_iso_dsig} with 
    disk parameters $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The real
    (imaginary) part of $\p\hat{\sigma}/\p\beta$ is shown as the solid
    (dotted) lines. 
    \label{domegadbeta}}  
\end{figure}

@@

\subsubsection{Dispersion relation with thermal
  relaxation}\label{disp_relax}
To obtain a dispersion relation similar to Eq. \ref{sig2_iso} for
isothermal perturbations, we first write 
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2$ to remain finite. Then assuming $g(\zhat)$ is a
polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}
We choose 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right),  
\end{align} 
by considering the case $q=0$ and $\beta\to\infty$ as examined in
\cite{lubow93}.

We now have an equation in the same form as that for isothermal
perturbations (Eq. \ref{iso_ode3}):
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp0}
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
% Eq. \ref{relax_disp} contrasts to Eq. \ref{adia_disp}, which is
% quadratic in $\hat{\sigma}$ for adiabatic perturbations.

It is simplest to solve Eq. \ref{relax_disp} numerically. In
Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
($M=0$) VSI mode as a function of $\khat$ for a range of
dimensionless thermal relaxation timescales $\beta$, in a disk with
$\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. Increasing $\beta$ even
slightly above zero significantly reduces the growth rates for
$\khat\gg 1$. (There is little effect at small $\khat$ but this case is
not of interest because the local model breaks down.) Thus, 
a growing fundamental VSI mode exists only for nearly
isothermal perturbations ($\beta\ll 1$).  The plot is also consistent with our previous
discussion that there exists an intermediate wavenumber for which
finite thermal relaxation has maximal effect (for small $\beta$). 
%$k_x\to0$ or $k_x\to\infty$ as $\beta\to0$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm 0cm]{figures/rate_theory_grow_relax}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_grow_relax2}
  \caption{Growth rate of the fundamental VSI mode in a disk with
    dimensionless thermal relaxation timescales
    $\beta\in[10^{-3},1]$.  Other disk parameters are
    $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The growth rates are
    calculated from the dispersion relation Eq. \ref{relax_disp}. 
    \label{relax_disp_fig}}  
\end{figure}

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} in
the limit $\khat\to\infty$ we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
We seek solutions with
\begin{align}
  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate.

In Fig. \ref{relax_bcrit} we plot the critical thermal relaxation
timescale obtained from solving Eq. \ref{relax_disp_fund} numerically 
for a disk with $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, for which
we have $\beta_\mathrm{crit}=0.125$ from Eq. \ref{iso_vsi_cond}. Our
upper estimate agrees well with the numerical solution for $\khat\gtrsim
10$. From our discussion for isothermal perturbations
(\S\ref{iso_discuss}), we expect $\beta<\beta_\mathrm{crit}$ to be a
sufficient condition for the VSI.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/bcrit_theory}
  \caption{Condition on the thermal relaxation timescale $\beta$ for
    the existence of unstable fundamental VSI mode in a vertically
    isothermal disk with $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.  
    The dotted line denotes the asymptotic value
    $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$ for large $\khat$. 
    \label{relax_bcrit}}  
\end{figure}

@@

\subsection{Spurious growth of adiabatic perturbations in the local
  model}\label{analytic_adia} 
For adiabatic perturbations ($\gamma>1$, $\beta\to\infty$) we may
simply use the Soilberg-Hoiland critera to assess stability
(\S\ref{solberg}). Nevertheless, we briefly consider this case in the
local framework to demonstrate its limitations. For $\beta\to\infty$
the parameter $\chi = 1/\gamma$ and Eq. \ref{vertiso_gov_nondim} can
be written as

\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat \notag\\
  &=  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
  &\phantom{=} +
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat\notag\\
  &\phantom{=} + \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear.

@@

However, in the presence of vertical shear $q\neq0$,
Eq. \ref{adia_integral} suggests $\hat{\sigma}$ is complex for real
$\khat$, which implies the possibility of instability, regardless of
how much $\gamma$ deviates from unity. This is contradicts the second 
Solberg-Hoiland criterion (Eq. \ref{solberg2}) and is a result of the
radially-local approximation (Appendix \ref{global_corr}). To 
avoid this, we must restrict our attention to short thermal relaxation
timescales ($\beta\lesssim1$) and/or neutrally-stratified disks
($\gamma=\Gamma$) when allowing vertical shear.

>>>>

Using the expressions for $\delta\chi$ and $\hat{\sigma}$ we can solve
for $\delta\hat{\sigma}$. This gives 
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
      q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
  \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
      \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by small finite cooling is ineffective at very large or
very small scales. 
% We plot Eq. \ref{nearly_iso_dsig} in
% Fig. \ref{domegadbeta} with 
% $\epsilon=0.05$, $q=-1.0$ and $\gamma=1.4$. For this disk model the
% optimum wavenumber for stabilization is $\hat{k}\sim 5$---$6$.

@@

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/domegadbeta}
%   \caption{Rate of change of the fundamental VSI eigenfrequency
%     $\sigma$ with respect to the introduction of a small but finite
%     cooling time. The plotted quantity is $\p\hat{\sigma}/\p\beta$ at $\beta=0$, evaluated using
%     Eq. \ref{nearly_iso_dsig} with 
%     disk parameters $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The real
%     (imaginary) part of $\p\hat{\sigma}/\p\beta$ is shown as the solid
%     (dotted) lines. 
%     \label{domegadbeta}}  
% \end{figure}

@@

\subsubsection{Dispersion relation with thermal
  relaxation}\label{disp_relax}
To obtain a dispersion relation similar to that for 
isothermal perturbations (Eq. \ref{sig2_iso}), we first write 
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. Inserting
Eq. \ref{adia_ansatz} into Eq. \ref{nearly_iso_explicit} gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
    \alpha\right)g
  +\left(\alpha^2 - \alpha A - C\right)\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$
vanish, and impose the vertical kinetic energy density
$\rho|\delta v_z|^2$ to remain finite. Then assuming $g(\zhat)$ is a
polynomial, we require  
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}
We choose 
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right),  
\end{align} 
by considering the case $q=0$ and $\beta\to\infty$ as examined in
\cite{lubow93}.

We now have an equation in the same form as that for isothermal
perturbations (Eq. \ref{iso_ode3}):
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp0}
\end{align}
% We remark that for $q=0$ and $\beta\to\infty$, Eq. \ref{adia_disp0} is
% the same as Eq. 54 in \cite{lubow93} when the low-frequency, Keplerian
% approximation is made in the latter. 
% Eq. \ref{nearly_iso_explicit} is same form
% as the governing equation for the adiabatic case
% (Eq. \ref{adia_iso4}). However, the coefficients $\widetilde{B}$ and $\widetilde{C}$
% have a complicated dependence on $\hat{\sigma}$. 
% Applying the same solution method as in \S\ref{adia_explicit}, we 
% find the adiabatic dispersion relation, 
This is a $6^\mathrm{th}$ order 
polynomial in $\hat{\sigma}$:  
\begin{align}
  0 = \sum_{l=0}^{6}c_l\hat{\sigma}^l,\label{relax_disp}
\end{align}
where the coefficients $c_l$ are given in Appendix \ref{relax_coeff}.
This dispersion relation generally requires a numerical
solution. However, for the fundamental mode ($M=0$) some analytical
progress can be made, which we consider next.

% It is simplest to solve Eq. \ref{relax_disp} numerically. In
% Fig. \ref{relax_disp_fig} we plot the growth rate of the fundamental
% ($M=0$) VSI mode as a function of $\khat$ for a range of
% dimensionless thermal relaxation timescales $\beta$, in a disk with
% $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. Increasing $\beta$ even
% slightly above zero significantly reduces the growth rates for
% $\khat\gg 1$. (There is little effect at small $\khat$ but this case is
% not of interest because the local model breaks down.) Thus, 
% a growing fundamental VSI mode exists only for nearly
% isothermal perturbations ($\beta\ll 1$).  The plot is also consistent with our previous
% discussion that there exists an intermediate wavenumber for which
% finite thermal relaxation has maximal effect (for small $\beta$).

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm 0cm]{figures/rate_theory_grow_relax}
% %  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/rate_theory_grow_relax2}
%   \caption{Growth rate of the fundamental VSI mode in a disk with
%     dimensionless thermal relaxation timescales
%     $\beta\in[10^{-3},1]$.  Other disk parameters are
%     $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The growth rates are
%     calculated from the dispersion relation Eq. \ref{relax_disp}. 
%     \label{relax_disp_fig}}  
% \end{figure}

@@

Here we estimate the maximum thermal relaxation timescale
$\beta$ that allows the fundamental VSI for $\khat\gg1$.   
% Notice in Fig. \ref{relax_bcrit} the upper limit on $\beta$ for the
% fundamental isothermal VSI tends to a constant for large
% $\khat$. Using this clue, we can calculate this critical $\beta$ by
% considering $\khat\to\infty$. We proceed as follows. 
At marginal stability, the eigenfrequency $\hat{\sigma}=\hat{\omega}$
is real. The dispersion relation for the fundamental mode is then
\begin{align}\label{relax_disp_fund}
  0 = \sum_{l=1}^{6}c_l\hat{\omega}^l \quad \text{for $M=\hat{\nu}=0$}.  
\end{align} 
Taking the real and imaginary parts of Eq. \ref{relax_disp_fund} in
the limit $\khat\to\infty$ we find
\begin{align}
  0 =& \left[1 + \gamma\beta^2\left(1-\gamma\right)\right] + 2\epsilon q
  \gamma\beta (\hat{\omega}\khat) -  (\hat{\omega}\khat)^2 \notag\\
  &+ \beta^2\gamma^2\hat{\omega}^2 (\hat{\omega}\khat)^2,\label{relax_cond1}\\
  0=& \beta(1-\gamma)\khat^2 + \epsilon q \khat^2 (\hat{\omega}\khat)
  - 2\beta (\hat{\omega}\khat)^2 - \epsilon q \gamma^2\beta^2 (\hat{\omega}\khat)^3.\label{relax_cond2}
\end{align}
We seek solutions with
\begin{align}
  \khat \to \infty, \quad \hat{\omega}\to 0 ,\quad \hat{\omega} \khat\text{ finite}.
\end{align}
Then
\begin{align}
  \hat{\omega}\khat = \frac{(\gamma-1)\beta}{\epsilon q}
\end{align}
from Eq. \ref{relax_cond2}, which implies, from Eq. \ref{relax_cond1}
that
\begin{align}
  \beta = \frac{1}{(\gamma-1)}\left[\frac{1}{\left(\epsilon
        q\right)^2} - \frac{\gamma}{(\gamma-1)}\right]^{-1/2} 
\end{align}
at marginal stability for perturbations with large $\khat$. 
For a thin disk $\epsilon\ll 1$, we infer 
\begin{align}\label{iso_vsi_cond}
  \beta \lesssim \frac{|\epsilon q|}{\gamma-1} \equiv
  \beta_\mathrm{crit} 
\end{align}
is required for the fundamental VSI to operate. From our discussion
for isothermal perturbations (\S\ref{iso_discuss}), we expect
$\beta<\beta_\mathrm{crit}$ to be a sufficient condition for the
VSI. That is, when $\beta<\beta_\mathrm{crit}$, perturbations are
effectively isothermal. For a thin disk 
$\beta_\mathrm{crit}\ll 1$, which suggests that finite thermal
relaxation rapidly stabilizes the VSI.

% In Fig. \ref{relax_bcrit} we plot the critical thermal relaxation
% timescale obtained from solving Eq. \ref{relax_disp_fund} numerically 
% for a disk with $\epsilon = 0.05$, $q=-1$ and $\gamma=1.4$, for which
% we have $\beta_\mathrm{crit}=0.125$ from Eq. \ref{iso_vsi_cond}. Our
% upper estimate agrees well with the numerical solution for $\khat\gtrsim
% 10$.

% \begin{figure}
%   \includegraphics[width=\linewidth]{figures/bcrit_theory}
%   \caption{Condition on the thermal relaxation timescale $\beta$ for
%     the existence of unstable fundamental VSI mode in a vertically
%     isothermal disk with $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$.  
%     The dotted line denotes the asymptotic value
%     $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$ for large $\khat$. 
%     \label{relax_bcrit}}  
% \end{figure}

@@

@@
